---
title: "Practicum1.CS5200Su21"
output:
  pdf_document: default
---
Authors:
Sunjit Dhillon [dhillon.su@northeastern.edu] 
MManzur Morshed [morshed.mm@northeastern.edu]

Link to LucidChart Diagram: https://lucid.app/lucidchart/05c683b2-bfc9-456a-8236-e299310aca73/edit?invitationId=inv_cea08688-d283-4f6b-9021-73ff624cccb3

```{r}
# Load the required libraries
library(RMySQL)
library(dplyr)
library(ggplot2)
```

```{r}
# Settings
db_user <- 'root'
db_password <- 'sunjit22'
db_name <- 'Practicum'

db_host <- 'localhost' 
db_port <- 3306

# Connect to DB
mydb <-  dbConnect(MySQL(), user = db_user, password = db_password,
                 dbname = db_name, host = db_host, port = db_port)

```

```{r}
# Read data from .csv file into a dataframe
path <- "/Users/sunjitdhillon/Downloads"
fn <- "BirdStrikesData.csv"

fileName <- paste(path, fn, sep = "/")

df <- read.csv(fileName, header = TRUE, stringsAsFactors = FALSE)
```

```{sql connection=mydb}
set global local_infile=true;
```

```{r}
# Create a temporary dataframe and rename column names
temp <- df
temp <- rename(temp, record_id = Record.ID, 
               aircraft_type=Aircraft..Type, 
               airport_name=Airport..Name, 
               altitude_bin=Altitude.bin, 
               aircraft_make_model=Aircraft..Make.Model, 
               wildlife_number_struck=Wildlife..Number.struck,
               wildlife_number_struck_actual=Wildlife..Number.Struck.Actual,
               impact_to_flight=Effect..Impact.to.flight,
               flight_date=FlightDate,
               indicated_damage=Effect..Indicated.Damage,
               aircraft_number_of_engines=Aircraft..Number.of.engines.,
               aircraft_airline_operator=Aircraft..Airline.Operator,
               origin_state=Origin.State, 
               phase_of_flight=When..Phase.of.flight, 
               conditions_precipitation=Conditions..Precipitation, 
               conditions_sky=Conditions..Sky, 
               remains_collected=Remains.of.wildlife.collected.,
               remains_sent_to_smithsonian=Remains.of.wildlife.sent.to.Smithsonian, 
               is_aircraft_large=Is.Aircraft.Large., 
               species=Wildlife..Species, 
               size=Wildlife..Size, 
               pilot_warned_of_birds_or_wildlife=Pilot.warned.of.birds.or.wildlife., 
               total_cost_in_dollars=Cost..Total.., 
               feet_above_ground=Feet.above.ground,
               number_of_people_injured=Number.of.people.injured,
               remarks = Remarks
)
head(temp)
```

Data Cleaning:
The aircraft_number_of_engines refers to the number of engines in an aircraft,
which must be an integer value. Based on analysis of data, we found that value 
of aircraft_number_of_engines corresponding to record id 206990 is 'C'. 
For uniformity of data type, we assume that aircraft_number_of_engines
corresponding to aircraft_make_model = 'RKWLTRBO 690' is 2. 
```{r}
r<-temp[which(temp$aircraft_make_model=='RKWLTRBO 690'), ]
r <- r %>% select(record_id, aircraft_make_model, aircraft_number_of_engines)
r
```

```{r}
# Replace 'C' with 2 in temp dataframe.
temp["aircraft_number_of_engines"][temp["aircraft_number_of_engines"] == 
                                     'C'] <- 2
```


```{r}
# Create dataframe Aircraft_df to store all distinct entries of aircrafts

Aircraft_df <- select(temp, aircraft_make_model, aircraft_number_of_engines, 
                      aircraft_type, is_aircraft_large)
Aircraft_df <- distinct(Aircraft_df)
aircraft_id <- seq_len(nrow(Aircraft_df))
Aircraft_df <- cbind(aircraft_id, Aircraft_df)
```

```{sql connection=mydb}
DROP TABLE IF EXISTS Aircraft
```

```{sql connection=mydb}
CREATE TABLE Aircraft(
               aircraft_id INTEGER NOT NULL,
       aircraft_make_model TEXT,    
aircraft_number_of_engines INTEGER,
             aircraft_type TEXT, 
         is_aircraft_large TEXT,
CONSTRAINT ck_categorical_aircraft_size CHECK (is_aircraft_large IN ("Yes","No", null)),
PRIMARY KEY (aircraft_id)
)
```

```{r}
# Write data from dataframe Aircraft_df to table Aircraft

dbWriteTable(mydb, "Aircraft", Aircraft_df, append = TRUE, row.names = FALSE)
```

```{sql connection=mydb}
SELECT * FROM Aircraft LIMIT 10
```

```{r}
# Create dataframe Airport_df to store all distinct entries of airports.

Airport_df <- select(temp, airport_name, origin_state)
Airport_df <- distinct(Airport_df)
airport_id <- seq_len(nrow(Airport_df))
Airport_df <- cbind(airport_id, Airport_df)
```

```{sql connection=mydb}
DROP TABLE IF EXISTS Airport
```

```{sql connection=mydb}
CREATE TABLE Airport(
  airport_id INTEGER NOT NULL,
airport_name TEXT,
origin_state TEXT,
PRIMARY KEY(airport_id)
)
```

```{r}
# Write data from dataframe Airport_df to table Airport
dbWriteTable(mydb, "Airport", Airport_df, append = TRUE, row.names = FALSE)
```
```{sql connection=mydb}
SELECT * FROM Airport LIMIT 10
```


```{r}
# Create dataframe Wildlife_df to store all distinct entries of Wildlife species.

Wildlife_df <- select(temp, species, size)
Wildlife_df <- distinct(Wildlife_df)
wildlife_id <- seq_len(nrow(Wildlife_df))
Wildlife_df <- cbind(wildlife_id, Wildlife_df)
```

```{sql connection=mydb}
DROP TABLE IF EXISTS Wildlife
```

```{sql connection=mydb}
CREATE TABLE Wildlife(
wildlife_id INTEGER NOT NULL,
    species TEXT,
       size TEXT,
CONSTRAINT ck_categorical_size CHECK (size IN ("Small","Medium","Large", null)),
PRIMARY KEY (wildlife_id)
)
```

```{r}
# Write data from dataframe Wildlife_df to table Wildlife

dbWriteTable(mydb, "Wildlife", Wildlife_df, append = TRUE, row.names = FALSE)
```

```{sql connection=mydb}
SELECT * FROM Wildlife LIMIT 10
```


```{r}
# Add aircraft_id column to temp dataframe

for(i in 1:dim(temp)[1]) {
  for(j in 1:dim(Aircraft_df)[1]) {
    
    # Compare values of aircraft_make_model, aircraft_number_of_engines, 
    # aircraft_type, is_aircraft_large in Aircraft_df and temp dataframe
    if (temp$aircraft_make_model[i]==Aircraft_df$aircraft_make_model[j] 
        & temp$aircraft_number_of_engines[i]==Aircraft_df$aircraft_number_of_engines[j] 
        & temp$aircraft_type[i]==Aircraft_df$aircraft_type[j] 
        & temp$is_aircraft_large[i]==Aircraft_df$is_aircraft_large[j]) {
      
      temp$aircraft_id[i] <- j
      break
    }
  }
}
```

```{r}
# Add airport_id column to temp dataframe

for(i in 1:dim(temp)[1]) {
  for(j in 1:dim(Airport_df)[1]) {
    
    # Compare values of airport_name, origin_state in Airport_df & temp dataframe
    if (temp$airport_name[i]==Airport_df$airport_name[j] &
        temp$origin_state[i]==Airport_df$origin_state[j]) {
      temp$airport_id[i] <- j
      break
    }
  }
}
```

```{r}
# Create dataframe Flight_Detail_df
Flight_Detail_df <- select(temp, record_id, aircraft_id, airport_id, 
                           flight_date, aircraft_airline_operator, 
                           pilot_warned_of_birds_or_wildlife)
# Change the flight_date format
Flight_Detail_df$flight_date <- as.Date(Flight_Detail_df$flight_date,"%m/%d/%Y")
# Assume missing values to be a default date '1900-01-01'
Flight_Detail_df$flight_date[is.na(Flight_Detail_df$flight_date)] <- '1900-01-01'
```

```{sql connection=mydb}
DROP TABLE IF EXISTS Flight_Detail
```

```{sql connection=mydb}
CREATE TABLE Flight_Detail(
                        record_id INTEGER NOT NULL,
                      aircraft_id INTEGER NOT NULL,
                       airport_id INTEGER NOT NULL,
                      flight_date DATE DEFAULT(DATE_FORMAT('%Y-%m-%d','1900-01-01')) NOT NULL,
        aircraft_airline_operator TEXT,
pilot_warned_of_birds_or_wildlife TEXT,
CONSTRAINT ck_categorical_pilot_warned CHECK (pilot_warned_of_birds_or_wildlife
IN ("Y","N", null)),
PRIMARY KEY (record_id),
FOREIGN KEY (aircraft_id) REFERENCES Aircraft(aircraft_id) ON DELETE CASCADE,
FOREIGN KEY (airport_id) REFERENCES Airport(airport_id) ON DELETE CASCADE
)
```

```{r}
# Write data from dataframe Flight_Detail_df to table Flight_Detail

dbWriteTable(mydb, "Flight_Detail", Flight_Detail_df, append = TRUE, 
             row.names = FALSE)
```
```{sql connection=mydb}
SELECT * FROM Flight_Detail LIMIT 10
```

```{r}
# Create dataframe Strike_Impact_df
Strike_Impact_df <- select(temp, record_id, impact_to_flight, indicated_damage,
                           number_of_people_injured, total_cost_in_dollars,
                           remarks)

```

```{sql connection=mydb}
DROP TABLE IF EXISTS Strike_Impact
```

```{sql connection=mydb}
CREATE TABLE Strike_Impact(
               record_id INTEGER NOT NULL,
        impact_to_flight TEXT,
        indicated_damage TEXT,
number_of_people_injured INTEGER,     
   total_cost_in_dollars INTEGER,
                 remarks TEXT,
CONSTRAINT ck_categorical_impact CHECK (impact_to_flight IN ("Aborted Take-off",
"Engine Shut Down","None", "Other","Precautionary Landing", null)),
CONSTRAINT ck_categorical_damage CHECK (indicated_damage IN ("Caused damage",
"No damage", null)),
PRIMARY KEY (record_id),
FOREIGN KEY (record_id) REFERENCES Flight_Detail(record_id) ON DELETE CASCADE
)
```

```{r}
# Write data from dataframe Strike_Impact_df to table Strike_Impact
dbWriteTable(mydb, "Strike_Impact", Strike_Impact_df, append = TRUE,
             row.names = FALSE)
```

```{sql connection=mydb}
SELECT * FROM Strike_Impact LIMIT 10
```

```{r}
# Create dataframe Strike_Condition_df
Strike_Condition_df <- select(temp, record_id, altitude_bin, feet_above_ground,
                              conditions_sky, conditions_precipitation, 
                              phase_of_flight)
```

```{sql connection=mydb}
DROP TABLE IF EXISTS Strike_Condition
```

```{sql connection=mydb}
CREATE TABLE Strike_Condition(
               record_id INTEGER NOT NULL,
            altitude_bin TEXT,
       feet_above_ground INTEGER,
          conditions_sky TEXT,
conditions_precipitation TEXT,
         phase_of_flight TEXT,
CONSTRAINT ck_categorical_altitude_bin CHECK (altitude_bin IN ("< 1000 ft","> 1000 ft", null)),
CONSTRAINT ck_categorical_conditions_sky CHECK (conditions_sky IN ("No Cloud","Some Cloud","Overcast", null)),
CONSTRAINT ck_categorical_phase CHECK (phase_of_flight IN ("Approach","Climb", "Descent", "Landing roll", "Take-off run", "Taxi", "Parked", null)),
PRIMARY KEY (record_id),
FOREIGN KEY (record_id) REFERENCES Flight_Detail(record_id) ON DELETE CASCADE
)
```

```{r}
# Write data from dataframe Strike_Condition_df to table Strike_Condition

dbWriteTable(mydb, "Strike_Condition", Strike_Condition_df, append = TRUE, 
             row.names = FALSE)
```

```{sql connection=mydb}
SELECT * FROM Strike_Condition LIMIT 10
```

```{r}
# Add wildlife_id column to temp dataframe
for(i in 1:dim(temp)[1]) {
  for(j in 1:dim(Wildlife_df)[1]) {
    
    # Compare values of species and size in Wildlife_df & temp dataframe
    if (temp$species[i]==Wildlife_df$species[j] & temp$size[i]==Wildlife_df$size[j]) {
      temp$wildlife_id[i] <- j
      break
    }
  }
}
```

```{r}
# Create dataframe Wildlife_Strike_df

Wildlife_Strike_df <- select(temp, record_id, wildlife_id, wildlife_number_struck, 
                             wildlife_number_struck_actual, remains_collected, 
                             remains_sent_to_smithsonian)
```

```{sql connection=mydb}
DROP TABLE IF EXISTS Wildlife_Strike
```

```{sql connection=mydb}
CREATE TABLE Wildlife_Strike(
                    record_id INTEGER NOT NULL,
                  wildlife_id INTEGER NOT NULL,
       wildlife_number_struck TEXT,
wildlife_number_struck_actual INTEGER,
            remains_collected TEXT,
  remains_sent_to_smithsonian TEXT,
CONSTRAINT ck_categorical_number_struck CHECK (wildlife_number_struck IN ("1","2 to 10","11 to 100","Over 100",null)),
CONSTRAINT ck_categorical_remains_collected CHECK (remains_collected IN ("TRUE","FALSE", null)),
CONSTRAINT ck_categorical_remains_sent_to_smithsonian CHECK (remains_sent_to_smithsonian IN ("TRUE","FALSE", null)),
PRIMARY KEY (record_id),
FOREIGN KEY (wildlife_id) REFERENCES wildlife(wildlife_id) ON DELETE CASCADE,
FOREIGN KEY (record_id) REFERENCES Flight_Detail(record_id) ON DELETE CASCADE
)
```

```{r}
# Write data from dataframe Wildlife_Strike_df to table Wildlife_Strike

dbWriteTable(mydb, "Wildlife_Strike", Wildlife_Strike_df, append = TRUE, 
             row.names = FALSE)
```
```{sql connection=mydb}
SELECT * FROM Wildlife_Strike LIMIT 10
```


Ques 4.
```{sql connection=mydb}
SELECT aircraft_airline_operator, COUNT(DISTINCT record_id) AS count_bird_strikes
FROM Flight_Detail
WHERE record_id IN (SELECT record_id FROM Strike_Condition WHERE phase_of_flight IN ("Take-off run", "Climb"))
GROUP BY aircraft_airline_operator;
```

Ques 5.
```{sql connection=mydb}
SELECT airport_name, COUNT(f.record_id) as count
FROM Flight_Detail AS f
NATURAL JOIN Airport AS a
GROUP BY a.airport_name
HAVING count = (
SELECT MAX(x.count) FROM
(SELECT a.airport_name AS airport_name, count(f.record_id) as count
FROM Flight_Detail AS f
NATURAL JOIN Airport AS a
GROUP BY a.airport_name) x)
```

Ques 6.
```{sql connection=mydb}
SELECT EXTRACT(YEAR FROM flight_date) AS year, COUNT(record_id) AS count_bird_strikes
FROM Flight_Detail
GROUP BY year
ORDER BY year 
```

Ques 7.
```{r}
# Create a dataframe containing counts of bird strike incidents grouped by year and phase of flight

sqlCmd = "SELECT EXTRACT(YEAR FROM f.flight_date) AS year, s.phase_of_flight, COUNT(*) AS count
FROM Flight_Detail AS f
NATURAL JOIN Strike_Condition AS s
GROUP BY year, s.phase_of_flight
HAVING year >= 2008 AND  year <= 2011 
AND phase_of_flight IN ('Take-off run', 'Climb', 'Descent', 'Approach', 'Landing roll')"

df = dbGetQuery(mydb, sqlCmd)
df
```
```{r}
# Group the flight phases into 'Take-off/Climbing' and 'Descent/Approach/Landing'

 for(i in 1:dim(df)[1]) {
    if (df$phase_of_flight[i]=='Take-off run' ||  df$phase_of_flight[i]=='Climb') {
      df$phase_of_flight[i] <- 'Take-off/Climbing'
    } 
   
   if (df$phase_of_flight[i]=='Approach' ||  df$phase_of_flight[i]=='Landing Roll' 
       || df$phase_of_flight[i]=='Descent') {
      df$phase_of_flight[i] <- 'Descent/Approach/Landing'

    }
}
df
```

```{r}
# Group the bird strike incidents by their total sum per year (grouped by flight phase)

df2 <-df %>%
  group_by(year, phase_of_flight) %>%
  summarise(count=sum(count))
df2
```

```{r}
# Plot the dataframe df2 to form a grouped bar chart

ggplot(df2, aes(year, count, fill = phase_of_flight)) + 
  geom_bar(stat="identity", position = "dodge") + 
  scale_fill_brewer("Flight Phase", palette = "Set1") +
   labs(y="Count", x = "Year") +
ggtitle("Bird Strikes / Year") +
  theme(plot.title = element_text(hjust = 0.5))
```

Ques 8.
```{sql connection=mydb}
DROP PROCEDURE IF EXISTS Delete_Flight_Detail 
```

// The procedure Delete_Flight_Detail deletes bird strike incident record
// corresponding to the record id entered as a parameter
```{sql connection=mydb}
CREATE PROCEDURE Delete_Flight_Detail(IN id_to_delete INTEGER)
    BEGIN
      DELETE FROM Flight_Detail
      WHERE record_id=id_to_delete;
    END
```

// Before calling the procedure:
```{sql connection=mydb}
SELECT * FROM Flight_Detail where record_id = 1195
```

// Call the procedure
```{sql connection=mydb}
CALL Delete_Flight_Detail(1195)
```

// After calling the procedure, the record corresponding to record id 1195 has 
// been deleted
```{sql connection=mydb}
SELECT * FROM Flight_Detail where record_id = 1195
```
```{r}
dbDisconnect(mydb)
```
